---
title: "U.S. Health Insurance Coverage Report"
---


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(forcats)
library(scales)
library(ggrepel)
library(patchwork)
library(tibble)
library(janitor)
library(plotly)
library(tidyverse)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40"),
      axis.title    = element_text(size = 10),
      legend.position = "bottom"
    )
)
```



Introduction:

Healthcare coverage in the U.S. is something people talk about a lot, but it’s actually kind of hard to get a clear picture of how it’s distributed and what really affects it. We were especially interested in uninsured rates, since that directly reflects how many people are not covered at all.
So our main goal here is to first understand what insurance coverage looks like overall in the U.S., and then look at what factors are related to people being uninsured. In particular, we want to see whether programs like Medicaid seem to play a role.



```{r, echo=FALSE, message=FALSE, warning=FALSE}
#this chunk creates the inital bar chart

num_insurance <- read.csv("data/raw/num_insurance.csv")

# Sum all states
us_totals <- num_insurance %>%
  summarise(across(-c(Location, Total), sum))

# Convert to long format
us_long <- us_totals %>%
  pivot_longer(
    cols = everything(),
    names_to = "Coverage",
    values_to = "People"
  )

us_prop <- us_long %>%
  arrange(desc(Coverage)) %>%
  mutate(
    People = People / sum(People),   # convert to proportions
    pos = cumsum(People) - People / 2,
    vjust_val = ifelse(row_number() %% 2 == 0, 4.5, -3.5),
  )

ggplot(us_prop, aes(x = "", y = People, fill = Coverage)) +
  geom_bar(stat = "identity", width = 0.4) +
  geom_segment(
    aes(
      x    = 1,
      xend = 1 + vjust_val * -0.08,
      y    = pos,
      yend = pos
    ),
    color = "black",
    linewidth = 0.4
  ) +
  geom_text(
    aes(y = pos, label = Coverage, vjust = vjust_val),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_flip() +
  labs(
    title = "U.S. Health Insurance Coverage Totals",
    x = "",
    y = "Coverage Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    aspect.ratio = 0.15
  )
```

Overall Coverage in the US:

We start with a general breakdown of insurance coverage across the U.S.
From the chart, employer-based insurance clearly makes up the largest share by far. This isn’t too surprising, since many people get insurance through their jobs. Medicaid and Medicare also take up a significant portion, while non-group and military coverage are smaller.
Even though uninsured is not the largest category, it’s still a noticeable portion of the population. That’s what we focus on more in the next steps.



```{r, echo=FALSE, message=FALSE, warning=FALSE}
raw_coverage <- read_csv("data/raw/Heathcare_Coverage_US.csv")

df <- raw_coverage %>%
  select(-Total) %>%
  filter(!is.na(Location), !str_starts(Location, "Notes")) %>%
  drop_na(Employer) %>%
  rename_with(~ c("location", "employer", "non_group", "medicaid",
                  "medicare", "military", "uninsured"), everything())

national <- df |> filter(location == "United States")
states   <- df |> filter(location != "United States")

corr_vals <- states %>%
  select(-location) %>%
  cor()

vars <- colnames(corr_vals)

cor_mat <- as.data.frame(corr_vals) %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "corr") %>%
  mutate(
    row = match(var1, vars),
    col = match(var2, vars)
  ) %>%
  filter(row >= col) %>%   # lower triangle
  mutate(
    var1 = factor(var1, levels = vars),
    var2 = factor(var2, levels = rev(vars))   # flips y-axis nicely
  )

ggplot(cor_mat, aes(var1, var2, fill = corr)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(corr, 2)), size = 3.5) +
  scale_fill_gradient2(
    low = "#E74C3C",
    mid = "white",
    high = "#2980B9",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "Lower Triangle Correlation Matrix",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

```
Next, we explore how different types of coverage are related to each other across states.
One pattern that stands out is the relationship between Medicaid and uninsured rates. The correlation is negative, meaning that states with higher Medicaid coverage tend to have fewer uninsured people.
There is also a strong negative relationship between employer-based insurance and Medicaid, suggesting that states rely more on one type of coverage than the other.
Overall this indicates that Medicaid plays an important role in reducing the number of uninsured individuals.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

cor_mat <- cor_mat %>%
  mutate(
    relation = case_when(
      corr > 0 ~ "increase",
      corr < 0 ~ "decrease",
      TRUE ~ "no change"
    ),
    
    tooltip = paste0(
      "An increase in <b>", var1,
      "</b> coverage is correlated with an <b>",
      relation,
      "</b> in <b>", var2,
      "</b> coverage",
      "<br><br>Correlation: ", round(corr, 2)
    )
  )

p <- ggplot(cor_mat, aes(x = var1, y = var2, fill = corr, text = tooltip)) +
  geom_tile(color = "white") +
  
  # keeps correlation numbers visible
  geom_text(aes(label = round(corr, 2)), size = 3.5) +
  
  scale_fill_gradient2(
    low = "#E74C3C",
    mid = "white",
    high = "#2980B9",
    midpoint = 0,
    limits = c(-1,1)
  ) +
  
  labs(
    title = "Interactive Correlation Matrix",
    x = NULL,
    y = NULL
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )

ggplotly(p, tooltip = "text")


```

To better understand these relationships, we also created an interactive version of the correlation matrix.
This allows us to hover over each pair of variables and see more detailed descriptions of how they are related. It makes it easier to explore patterns without needing to read every number carefully.
From this interactive view, the same pattern becomes even clearer: higher Medicaid coverage is consistently associated with lower uninsured rates across states. At the same time, employer-based coverage and Medicaid tend to move in opposite directions.
Overall, the interactive chart helps confirm our earlier finding that Medicaid plays a key role in reducing the uninsured population.



Conclusion:

Our analysis shows that while most people in the US have some form of health insurance, there is still a significant portion of the population that remains uninsured.From the correlation analysis, Medicaid stands out as the most important factor associated with lower uninsured rates. States with higher Medicaid coverage tend to have fewer uninsured individuals, suggesting that Medicaid plays a key role in providing access to healthcare.This also helps us think about the potential impact of Medicaid cuts. If Medicaid coverage is reduced, it is likely that the number of uninsured people will increase, especially in states that rely more heavily on it.
To summarize, our findings highlight how important government programs like Medicaid are in maintaining health insurance coverage and reducing gaps in access to care.


Project Video: 

{{< video Project.mp4 >}}

Project Slides: 

<iframe src="Slides.pdf" width="100%" height="600px"></iframe>

