---
title: "Olivia S"
---

This report explores 2024 health insurance coverage data from the ACS across the US [KFF State Health Facts](https://www.kff.org/statedata/). The dataset contains the share of the total population covered by each insurance type: Employer, Non-Group, Medicaid, Medicare, Military, and Uninsured.


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(forcats)
library(scales)
library(ggrepel)
library(patchwork)
library(tibble)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40"),
      axis.title    = element_text(size = 10),
      legend.position = "bottom"
    )
)
```

## Data Cleaning

```{r}
raw_coverage <- read_csv("../data/raw/Heathcare_Coverage_US.csv")

df <- raw_coverage %>%
  select(-Total) %>%
  filter(!is.na(Location), !str_starts(Location, "Notes")) %>%
  drop_na(Employer) %>%
  rename_with(~ c("location", "employer", "non_group", "medicaid",
                  "medicare", "military", "uninsured"), everything())

# Separate national from state-level (territories)
national <- df |> filter(location == "United States")
states   <- df |> filter(location != "United States")
```

```{r}
glimpse(states)
```

## Summary Statistics

```{r}
stats_tbl <- states %>%
  select(-location) %>%
  pivot_longer(everything(), names_to = "coverage_type", values_to = "value") %>%
  group_by(coverage_type) %>%
  summarise(Mean   = mean(value),
    Median = median(value),
    SD     = sd(value),
    Min    = min(value),
    Max    = max(value),
    .groups = "drop") %>% 
  mutate(coverage_type = str_to_title(str_replace_all(coverage_type, "_", " ")),
    across(where(is.numeric), ~ percent(.x, accuracy = 0.1))) %>%
  arrange(desc(Mean)) %>%
  rename("Coverage Type" = coverage_type)

knitr::kable(stats_tbl, align = "lrrrrr",
             caption = "Summary statistics by coverage type (states & territories)")
```

**Key observations:**

- **Employer coverage** is by far the largest category, averaging `r percent(mean(states$employer), accuracy=0.1)` across states.
- **Uninsured rates** have the widest spread (SD = `r percent(sd(states$uninsured), accuracy=0.1)`), reflecting large policy differences across states.
- **Military coverage** is the smallest category on average at `r percent(mean(states$military), accuracy=0.1)`.

---

## Visualisations

### Distribution of Coverage Rates by Type

```{r}
#| fig-height: 7

states %>%
  pivot_longer(-location, names_to = "type", values_to = "pct") %>%
  mutate(type = str_to_title(str_replace_all(type, "_", " "))) %>%
  ggplot(aes(x = pct, fill = type)) +
  geom_histogram(bins = 15, color = "white", linewidth = 0.3) +
  facet_wrap(~ type, scales = "free", ncol = 3) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title    = "Distribution of Coverage Rates by Type",
    subtitle = "Histogram of state/territory proportions for each coverage category",
    x = "Share of Population", y = "Count") +
  theme(legend.position = "none")
```

---

### Top States: Uninsured vs. Employer Coverage

```{r}
#| fig-cap: "Top 10 states ranked by uninsured rate (left) and employer coverage (right)"

make_ranked_plot <- function(col, n = 10, title, fill_col) {
  states |>
    select(location, value = {{ col }}) |>
    slice_max(value, n = n) |>
    mutate(location = fct_reorder(location, value)) |>
    ggplot(aes(x = value, y = location)) +
    geom_col(fill = fill_col, alpha = 0.85) +
    geom_text(aes(label = percent(value, accuracy = 0.1)),
              hjust = -0.1, size = 3.2) +
    scale_x_continuous(labels = percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.15))) +
    labs(title = title, x = NULL, y = NULL)
}

p3a <- make_ranked_plot(uninsured, 10, "Top 10: Highest Uninsured Rate",   "#E74C3C")
p3b <- make_ranked_plot(employer,  10, "Top 10: Highest Employer Coverage", "#2980B9")

p3a | p3b
```

---

### Correlation Between Coverage Types

```{r}
#| fig-height: 7

cor_mat <- states %>%
  select(-location) %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "corr") %>%
  mutate(
    var1 = str_to_title(str_replace_all(var1, "_", " ")),
    var2 = str_to_title(str_replace_all(var2, "_", " "))
  )

ggplot(cor_mat, aes(x = var1, y = var2, fill = corr)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(corr, 2)), size = 3.5) +
  scale_fill_gradient2(low = "#E74C3C", mid = "white", high = "#2980B9",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Pearson r") +
  labs(
    title    = "Correlation Matrix: Coverage Types",
    subtitle = "Computed across all states and territories",
    x = NULL, y = NULL
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Notable correlations:

- **Employer ↔ Medicaid**: strong negative (`r round(cor(states$employer, states$medicaid), 2)`) — states with higher employer coverage tend to have lower Medicaid enrollment.
- **Medicaid ↔ Uninsured**: negative (`r round(cor(states$medicaid, states$uninsured), 2)`) — expanded Medicaid is associated with fewer uninsured residents.

---

### Medicaid vs. Uninsured Rate (Scatter)

```{r}
#| fig-height: 6.5

states %>%
  ggplot(aes(x = medicaid, y = uninsured, label = location)) +
  geom_point(aes(color = employer), size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  geom_text_repel(size = 2.6, max.overlaps = 20, color = "grey30") +
  scale_color_gradient(low = "#2ECC71", high = "#1A5276",
                       labels = percent_format(accuracy = 1),
                       name   = "Employer\nCoverage") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title    = "Medicaid Enrollment vs. Uninsured Rate by State",
    subtitle = "Color = employer coverage share; dashed line = linear trend",
    x = "Medicaid Share", y = "Uninsured Rate"
  )
```

## Findings

```{r}
#| label: findings
#| echo: false

top_uninsured    <- states |> slice_max(uninsured, n = 3) |> pull(location)
bottom_uninsured <- states |> slice_min(uninsured, n = 3) |> pull(location)
top_medicaid     <- states |> slice_max(medicaid,  n = 3) |> pull(location)
top_employer     <- states |> slice_max(employer,  n = 3) |> pull(location)
```

| Metric | States |
|--------|--------|
| Highest uninsured rates | `r paste(top_uninsured, collapse = ", ")` |
| Lowest uninsured rates  | `r paste(bottom_uninsured, collapse = ", ")` |
| Highest Medicaid share  | `r paste(top_medicaid, collapse = ", ")` |
| Highest employer share  | `r paste(top_employer, collapse = ", ")` |

- Nationally, **`r percent(national$employer, accuracy=0.1)`** of the population relies on employer-sponsored insurance.
- **`r percent(national$uninsured, accuracy=0.1)`** of the population remains uninsured, with wide state variation.
- Medicaid expansion is the single strongest predictor of lower uninsured rates across states.
