---
title: "Jack Fang"
---

y
```{r}
library(janitor)
library(tidyverse)
library(ggplot2)
data <- read_csv("../data/raw/Heathcare_Coverage_US.csv")
data <- clean_names(data)
```
```{r}
summary(data)
```

```{r}
ggplot(data, aes(x = uninsured)) +
  geom_histogram(bins = 20, fill = "blue") +
  theme_minimal()
```
```{r}
data |> arrange(desc(uninsured)) |> select(location, uninsured) |> head(10)
```
```{r}
ggplot(data, aes(x = medicaid, y = uninsured)) +
  geom_point() +
  theme_minimal()
```
```{r}
ggplot(data, aes(x = employer, y = uninsured)) +
  geom_point() +
  theme_minimal()
```
```{r}
data_long <- data |>
  select(location, employer, medicaid, medicare, uninsured) |>
  pivot_longer(-location)

ggplot(data_long, aes(x = name, y = value)) +
  geom_boxplot() +
  theme_minimal()
```
Observation:

There are some noticeable differences in uninsured rates across states. Most states are between about 5% and 9%, but a few states have much higher values, with Texas being a clear outlier.
There seems to be a slight negative relationship between Medicaid coverage and uninsured rates, although the pattern is not very strong.
The relationship between employer-based coverage and uninsured rates is more clear. States with higher employer coverage tend to have lower uninsured rates.
Overall, both public programs like Medicaid and employer-based insurance appear to play a role in explaining differences in uninsured rates across states.
