---
title: "Masha A"
---

Your exploratory data analysis of the team datasets go here.

## Uninsured Rates for Adults Ages 19-64 by Sex

```{r}
library(tidyverse)
library(janitor)
library(gtsummary)
library(visdat)

uninsured_data <- read_csv("../data/raw/uninsured_sex_2024.csv", skip = 2) |> 
  clean_names()|>
  filter(!is.na(male),
         !location %in% c("Notes", "Sources", "Definitions", "Footnotes"),
         location != "United States")
```

## Plot 1: Overall distribution of uninsured rates (total) 

```{r}
ggplot(uninsured_data, aes(x = total)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Overall Distribution of Uninsured Rates Across States",
    subtitle = "Adults ages 19–64 | Source: KFF/ACS 2024",
    x = "Uninsured Rate",
    y = "Number of States"
  ) +
  theme_minimal()
```


## Plot 2: Distribution by sex

```{r}
uninsured_long <- uninsured_data |>
  pivot_longer(cols = c(male, female), names_to = "sex", values_to = "rate")

ggplot(uninsured_long, aes(x = rate, fill = sex)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("male" = "#2166ac", "female" = "#d6604d"),
                    labels = c("Male", "Female")) +
  labs(
    title = "Distribution of Uninsured Rates by Sex Across States",
    x = "Uninsured Rate",
    y = "Density",
    fill = "Sex"
  ) +
  theme_minimal()
```



## Plot 3: Gender gap for ALL states

```{r}
uninsured_data |>
  mutate(location = fct_reorder(location, total)) |>
  ggplot() +
  geom_segment(aes(x = female, xend = male,
                   y = location, yend = location),
               color = "grey70", linewidth = 0.8) +
  geom_point(aes(x = male, y = location), color = "#2166ac", size = 2) +
  geom_point(aes(x = female, y = location), color = "#d6604d", size = 2) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Uninsured Rates by Sex for All States",
    subtitle = "Blue = Male  |  Red = Female  |  Ordered by total uninsured rate",
    x = "Uninsured Rate (Ages 19–64)",
    y = NULL,
    caption = "Source: KFF/ACS 2024"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 7))
```


## Initial EDA Observations

Gender Disparity - Initial density plots suggest that males generally have a higher uninsurance rate than females in the 19-64 age bracket. This may be linked to female eligibility for programs like Medicaid during pregnancy or different employment patterns

State Volatility - There is a range in rates. States that did not expand Medicaid likely show higher total uninsured percentages for both sexes compared to expansion states
