---
title: "Masha A"
---

Your exploratory data analysis of the team datasets go here.

## Uninsured Rates for Adults Ages 19-64 by Sex

```{r}
library(tidyverse)
library(janitor)
library(gtsummary)
library(visdat)

uninsured_data <- read_csv("../data/raw/uninsured_sex_2024.csv", skip = 2) |> 
  clean_names()
```


## Is there a gender gap in uninsurance rates?

```{r}
uninsured_long <- uninsured_data |>
  filter(location != "United States") |> 
  pivot_longer(cols = c(male, female), names_to = "sex", values_to = "rate")

ggplot(uninsured_long, aes(x = rate, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Uninsured Rates by Sex Across States",
       x = "Uninsured Rate (%)",
       y = "Density") +
  theme_minimal()
```



## Geographic Variation

```{r}
uninsured_data |>
  filter(location != "United States") |>
  mutate(gap = male - female) |>
  slice_max(gap, n = 10) |>
  ggplot(aes(x = reorder(location, gap), y = gap)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 States with Largest Gender Gap",
       subtitle = "States where men are significantly more likely to be uninsured than women",
       x = "State",
       y = "Difference (Male % - Female %)") +
  theme_minimal()
```


## Initial EDA Observations

Gender Disparity - Initial density plots suggest that males generally have a higher uninsurance rate than females in the 19-64 age bracket. This may be linked to female eligibility for programs like Medicaid during pregnancy or different employment patterns

State Volatility - There is a range in rates. States that did not expand Medicaid likely show higher total uninsured percentages for both sexes compared to expansion states
