---
title: "Viz 2"
format: html
---


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(forcats)
library(scales)
library(ggrepel)
library(patchwork)
library(tibble)
library(janitor)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "grey40"),
      axis.title    = element_text(size = 10),
      legend.position = "bottom"
    )
)
```

## Data Cleaning

```{r}
raw_coverage <- read_csv("../data/raw/Heathcare_Coverage_US.csv")

df <- raw_coverage %>%
  select(-Total) %>%
  filter(!is.na(Location), !str_starts(Location, "Notes")) %>%
  drop_na(Employer) %>%
  rename_with(~ c("location", "employer", "non_group", "medicaid",
                  "medicare", "military", "uninsured"), everything())

# Separate national from state-level (territories)
national <- df |> filter(location == "United States")
states   <- df |> filter(location != "United States")
```

## Summary Statistics

```{r}
stats_tbl <- states %>%
  select(-location) %>%
  pivot_longer(everything(), names_to = "coverage_type", values_to = "value") %>%
  group_by(coverage_type) %>%
  summarise(Mean   = mean(value),
    Median = median(value),
    SD     = sd(value),
    Min    = min(value),
    Max    = max(value),
    .groups = "drop") %>% 
  mutate(coverage_type = str_to_title(str_replace_all(coverage_type, "_", " ")),
    across(where(is.numeric), ~ percent(.x, accuracy = 0.1))) %>%
  arrange(desc(Mean)) %>%
  rename("Coverage Type" = coverage_type)

knitr::kable(stats_tbl, align = "lrrrrr",
             caption = "Summary statistics by coverage type (states & territories)")
```

**Key observations:**

- Employer coverage is by far the largest category, averaging `r percent(mean(states$employer), accuracy=0.1)` across states.
- Uninsured rates have the widest spread (SD = `r percent(sd(states$uninsured), accuracy=0.1)`), reflecting large policy differences across states.
- Military coverage is the smallest category on average at `r percent(mean(states$military), accuracy=0.1)`.

---

## Visualisations
```{r}
#| fig-height: 7

cor_mat <- states %>%
  select(-location) %>%
  cor() %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "corr") %>%
  mutate(
    var1 = str_to_title(str_replace_all(var1, "_", " ")),
    var2 = str_to_title(str_replace_all(var2, "_", " "))
  )

ggplot(cor_mat, aes(x = var1, y = var2, fill = corr)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(corr, 2)), size = 3.5) +
  scale_fill_gradient2(low = "#E74C3C", mid = "white", high = "#2980B9",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Pearson r") +
  labs(
    title    = "Correlation Matrix: Coverage Types",
    subtitle = "Computed across all states and territories",
    x = NULL, y = NULL
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

